Structuring Theories with Implicit Morphisms

### 2018-09-26 implicit.tex

\subsection{Motivation}

@@

\subsection{Definition}

\paragraph{Syntax}
We have already indicated that we will consider all identities and inclusions as implicit.
It remains to give users the option introduce additional implicit morphisms.
For that purpose, we allow two \mmt declarations to carry an attribute: 
\begin{grammar}
MDec   & Att\; m:T\to T=\{Ass,\ldots,Ass\}   & \text{attributed atomic morphism} \\
Dec    & Att\; \icl{S}                       & \text{attributed inclusion}\\
Att    & \keyword{implicit} \alt \keyword{invertible} & \text{attributes}
\end{grammar}

Every morphism with the attribute \keyword{implicit} occurs in the implicit-diagram.
(Of course, this is redundant for includes.)
Every morphism with the attribute \keyword{invertible} is also considered implicit, and additionally so is its inverse morphism.

We do not even have to provide a name for the inverse morphism: It is uniquely determined (up to $\equiv$) anyway, and it will be used only implicitly.
We only have to make sure that the inverse actually exists and that it can be found automatically.
This is not decidable in general, but it is decidable in many important special cases, in particular, renamings and definitional extensions.

\paragraph{Semantics}
Well-formedness and semantics are defined in parallel.

\begin{definition}[Implicit-Diagram]
Given a diagram $D$, its implicit-diagram $D^i$ contains all theories and the following morphisms:
\begin{compactitem}
\item all identities and inclusions,
\item all compositions of implicit morphisms,
\item every morphism with the \keyword{implicit} attribute,
\item for each morphism with the \keyword{invertible} attribute, that morphism and its inverse. 
\end{compactitem}
\end{definition}

The only requirement for well-formedness is that the implicit-diagram always commutes.
However, in general, commutativity may be undecidable.
Similarly, the existence of an inverse morphism may be undecidable.
Therefore, we use sound criteria for commutativity and invertibility as defined in Sect.~\ref{sec:commute} and \ref{sec:inverse}.

\begin{definition}[Well-Formedness]
Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.

Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
\end{definition}

Finally, we modify the definition of $\Exp{T}$ and $M(-)$ to add all implicitly imported identifiers:

@@

\subsection{Applications}

Canonical isomorphisms.

User-declared atomic morphisms.

\paragraph{Extending Theories from the Outside}

\paragraph{Transparent Refactoring}
A major drawback of OpenMath/\mmt-style qualified identifiers is that it can preclude transparent refactoring.
For example, consider a theory $t=\{\icl{r},\icl{s}\}$ and assume we want to move a symbol $n$ from $r$ to $s$.
This requires changing any qualified reference to $r?n$ (in any theory including $t$) to $s?n$.
That can be very costly, especially in large libraries where the refactorer might not even know of or have write access to all theories importing $t$.

With implicit morphisms, we rename $s$ to $s'$, move $n$ from $s'$ to $r$, and recover $s$ by $s=\{\icl{s'},n\}$.
Now we can change define $t=\{\icl{r},\icl{s'}\}$ and give an implicit morphism from $s$ to $t$, which maps $s?n$ to $r?n$.
All existing references to $s?n$ stay well-formed (and serve as aliases for $r?n$) so that no further changes in theories importing $t$ are needed.

>>>>

\subsection{Overview}

@@

\subsection{Syntax}

We introduce the family of sets $\IMo{S}{T}$ as a subset of $\Mo{S}{T}$, holding the \textit{implicit} morphisms.
The intuition is that $\Thy$ and $\IMo{S}{T}$ (up to equality of morphisms) form a thin broad subcategory of $\Thy$ and $\Mo{S}{T}$.

It remains to define which morphisms are implicit.
For that purpose, we allow \mmt declarations to carry \textit{attributes}.
Precisely, we add the following productions
\begin{grammar}
MDec   & Att\; MDec  & \text{attributed morphism} \\
Dec    & Att\; Dec   & \text{attributed declaration}\\
Att    & \keyword{implicit} \alt \ldots & \text{attributes}
\end{grammar}

The set of attributes is itself extensible, and the above grammar only list the one that we use to get started.
Additional attributes may be added with the condition that any additional attribute increases the set of implicit morphisms.

Specifically, the set $\IMo{S}{T}\sq\Mo{S}{T}$ of \textbf{implicit} morphisms contains the following elements:
\begin{itemize}
 \item all identity morphisms $\id{T}$,
 \item all compositions $M;N$ of implicit morphisms $M$ and $N$,
 \item all declared morphisms $m:S\to T=\{\sigma\}$ whose declaration carries the attribute $\keyword{implicit}$.
\end{itemize}

\subsection{Semantics}

We only have to make two minor changes to the semantics to accommodate implicit morphisms.
The first change govern how we obtain implicit morphisms, the second one how we use them.

\paragraph{Obtaining}
We define well-formedness for our extended syntax.
We define that an attribute to a declaration is only well-formed if the category of implicit morphisms remains thin, i.e., for any two theories, there may be only one (up to equality) implicit morphism between them.
In other the diagram containing only the implicit morphisms must commute.

Depending on what modularity principles we add, equality of morphisms may or may not be decidable.
Therefore, in the general case, implementations may have to employ sound but incomplete criteria for commutativity (see Sect.~\ref{sec:commute}).

%\begin{definition}[Well-Formedness]
%Consider a well-formed diagram $D, m:S\to T=\{\sigma\}$.
%Let $D^a$ be $D, a\; m:S\to T=\{\sigma\}$.
%$D^{\keyword{implicit}}$ is well-formed if $D^i$ extended with an edge for $m$ commutes.
%$D^{\keyword{invertible}}$ is well-formed if $D^i$ extended with edges for $m$ and its inverse commutes.
%
%Correspondingly, consider a well-formed diagram $D$ of the form $D_0, T=\{\ldots,\icl{S},\ldots\}$.
%Let $D^a$ be $D_0, T=\{\ldots, a\; \icl{S},\ldots\}$.
%$D^{\keyword{invertible}}$ is well-formed if the inclusion has an *inverse.
%\end{definition}

\paragraph{Using}
We make a small modification to the rules in

@@

### 2018-09-26 mmt.tex

\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=p$, which establish theorem $F$ via proof $p$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T o:o'$ (typing) and $\der_T o\equiv o'$ (equality) for every theory $T$ and $o,o'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

@@

Both theories and morphisms are subject to typing conditions that are not essential for our purposes here.
We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.

@@

@@

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=o\}\]
\bigskip

The semantics serves two purposes:
firstly, it flattens every well-formed module into a flat one;
secondly, it determines which modules are well-formed at all.
It is desirable to define well-formedness independently of flattening.
However, in practice, it shows that implementations typically have to do almost the entire flattening to check well-formedness anyway.

Therefore, it is convenient to merge the two into one operation by simply making flattening a partial function: a module $X$ is \textbf{well-formed} iff $\flt{X}$ is defined.
\bigskip

@@

\subsection{Semantic Relations}

>>>>

\textbf{Flat theories} are lists of declarations $c[:E][=e]$ where $E$ and $e$ are optional expressions.
We write $\dom{T}$ for the set of constant identifiers $c$ in $\flt{T}$ and $\Exp{T}$ for the set of closed expressions using only the symbols $c\in\dom{T}$.
Constant declarations subsume virtually all basic declarations common in formal systems such as type/function/predicate symbols, axioms, theorems, inference rules, etc.
In particular, theorems can be represented via the propositions-as-types correspondence as declarations $c:F=P$, which establish theorem $F$ via proof $P$.
Similarly, \mmt expressions subsume virtually all objects common in formal systems such as terms, types, formulas, proofs.

Individual formal languages arise as fragments of \mmt: they single out the well-formed expressions by defining the two \mmt-\textbf{judgments} $\der_T e:e'$ (typing) and $\der_T e\equiv e'$ (equality) for every theory $T$ and $e,e'\in\Exp{T}$.
%This is already sufficient to define logical theorems as those formulas $F$ for which there is an object $p$ (the proof) such that $\der_T p:F$.
The details can be found in \cite{rabe:howto:14}.

@@

The declarations in theories and morphisms are subject to typing conditions that are not essential for our purposes here.
We only mention the main theorem that guarantees that well-typed morphisms preserve judgments, i.e., if $\vdash_S e:E$, then $\vdash_T M(e):M(E)$.
This includes the preservation of truth via the propositions-as-types principle where $E$ is a proposition and $e$ its proof.

@@

\paragraph{Well-Formedness}
The inference system to define well-formed modules, and expressions is not a primary interest of this paper.
Therefore, we only mention those aspects that are relevant later on.

However, one detail will be critical later on:
the inference rules for the well-formedness of expressions include the following base case for constants:
\[\rul{c:E[=e] \minn \flt{T}}{\vdash_T c:E} \tb \rul{c[:E]=e \minn \flt{T}}{\vdash_T c=e}\]
Correspondingly, the definition of the homomorphic extension of a morphism includes the following base case for constants:
\[M(c) = e \mwhere c:= e\in\flt{M}\]

These base cases introduce a mutual recursion between well-formedness and flattening:
the well-formedness of a declaration in a theory depends on the flattening of all preceding declarations;
and the well-formedness of an assignment in a morphism depends on the homomorphic extension of the morphism obtained by flattening of all preceding assignments.
Vice versa, well-formedness is a precondition for defining the flattening.
It is desirable to define well-formedness independently of flattening.
But the mutual recursion makes sense from an implementation perspective: typical tools for formal systems will first parse, check, and flatten the first declaration, then continue with the next declaration.
In fact, often the checking of well-formedness is as expensive as flattening anyway.

Therefore, we will make flattening a partial function, i.e., $\flt{X}$ is defined iff the module $X$ is well-formed.

%Consider any declaration $n[:E][=e]$ in the body of a theory $t$.
%It is well-formed if $n$ is a fresh name $E$ and $e$ are well-formed expressions over $t$.
%(Additionally, $E$ and $e$ must satisfy typing conditions that are not essential for our purposes.)
%Importantly, $E$ and $e$ must be from $\Exp{\flt{t}}$ using only constants that were introduced in preceding declarations.
%
%Similarly, consider an assignment $c:=e$ in a morphism $m$ from $S$ to $T$, where $c:E$ is a declaration is $S$.
%It is well-formed if $\vdash_T e:\flt{m}(E)$.

@@

Correspondingly, for a declared morphism $m:S\to T=\{\sigma\}$, we have $m\in\Mo{S}{T}$ and $\flt{m}$ is defined by induction on the assignments in $\sigma$:
 \[\flt{\cdot}=\es\]
 \[\flt{(\sigma,\;A)}=\flt{\sigma}\cup\fltd{\flt{\sigma}}{A}\]
with the trivial base case
 \[\fltd{\sigma}{c:=e}=\{c:=o\}\]

@@

\subsection{Semantic Properties}
